---
title: "Project"
output: html_document
date: "2024-05-31"
css: custom.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(recount3)
  library(reshape2)
  library(biomaRt)
  library(ggrepel)     # pca
  library(factoextra)  # k-means
  library(ggpubr)      # k-means
  library(randomForest)
  library(genefilter)  # lda
  library(MASS)        # lda
  library(pROC)
  library(caret)
  library(glmnet)
})

main_folder <- "/home/alice/ago2023/Desktop/unitn/NDA"
```

```{r}
metadata <- suppressMessages(read_delim(file.path(main_folder,"SraRunTable.txt"), delim = ",")) 

nevi <- metadata %>%
  filter(tissue == "benign melanocytic nevi") %>%
  filter(!Run=="SRR6916930") %>%
  filter(!Run=="SRR6916894") %>%
  filter(!Run=="SRR6916892")

set.seed(1)
melanoma <- metadata %>%
  filter(tissue == "primary melanoma") %>%
  filter(!Run=="SRR6916941") %>%
  sample_n(20)

groups <- rbind(melanoma,nevi) %>%
  dplyr::select(Run,tissue) %>%
  dplyr::rename("sample"=Run, "group"=tissue)
```

Download uniformly processed RNA-seq data from recount3
```{r}
## Get the expression data
proj_info <- subset(
    available_projects(),
    project == "SRP136742" & project_type == "data_sources"
    )
rse_gene_SRP136742 <- create_rse(proj_info)
rse_gene_SRP136742

assay(rse_gene_SRP136742, "counts") <- transform_counts(rse_gene_SRP136742)
ex <- rse_gene_SRP136742@assays@data@listData[["counts"]]
ex <- as.data.frame(ex) %>%
  dplyr::select(all_of(groups$sample))
```

Filtering lowly expressed genes
```{r}
suppressPackageStartupMessages({
  library(edgeR)
})

myCPM <- cpm(ex)
thresh <- myCPM > 0.3
# summary of how many TRUEs there are in each row
# There are 15023 genes that have TRUEs in all 12 samples.
table(rowSums(thresh))
# we would like to keep genes that have at least 2 TRUEs in each row of thresh
keep <- rowSums(thresh) >= 10
summary(keep)

mat = ex[keep,]

dim(ex)
# [1] 63856    40
dim(mat)
# [1] 24567    40
```

Let's have a look and see whether our threshold of 0.3 does indeed correspond to a count of about 10. We will look at the first sample.  
```{r}
plot(myCPM[,1],ex[,1])
# Let us limit the x and y-axis so we can actually look to see what is happening at the smaller counts
plot(myCPM[,1],ex[,1],ylim=c(0,50),xlim=c(0,3),
     xlab = "CPM",
     ylab = "Read counts")
# Add a vertical line at 0.3 CPM
abline(v=0.3)

png(file.path(main_folder,"CPM_threshold.png"), width = 2500, height = 1500, units="px",res=300)
plot(myCPM[,1],ex[,1],ylim=c(0,50),xlim=c(0,3),
     xlab = "CPM",
     ylab = "Read counts")
# Add a vertical line at 0.3 CPM
abline(v=0.3)
dev.off()
```

Histograms of counts per gene
An overview of the distribution of the genes counts for the sample SRR6916902
```{r}
a <- ggplot(mat) +
  geom_histogram(aes(x = SRR6916902), stat = "bin", bins = 200) +
  ggtitle("Count Distribution for SRR6916902 sample") +
  xlab("Raw expression counts") +
  ylab("Number of genes") +
  scale_x_continuous(labels=function(x) format(x, big.mark = "", scientific = FALSE))
a

b <- ggplot(mat) +
  geom_histogram(aes(x = SRR6916902), stat = "bin", bins = 200) +
  xlab("Raw expression counts (truncated axis)") +
  ylab("Number of genes") +
  scale_y_continuous(limits=c(0,2500)) +
  scale_x_continuous(labels=function(x) format(x, big.mark = "", scientific = FALSE), 
                     limits=c(0,2000))
b

c <- ggplot(log2(mat + 1)) +
  geom_histogram(aes(x = SRR6916902), stat = "bin", bins = 200) +
  ggtitle("Log2-transformed Count Distribution for SRR6916902 sample") +
  xlab("Log2(counts+1)") +
  ylab("Number of genes") 
c


png(file.path(main_folder,"hist_distribution_a.png"), width = 2500, height = 1500, units="px",res=300)
a
dev.off()

png(file.path(main_folder,"hist_distribution_b.png"), width = 2500, height = 1500, units="px",res=300)
b
dev.off()

png(file.path(main_folder,"hist_distribution_c.png"), width = 2500, height = 1500, units="px",res=300)
c
dev.off()
```


Boxplots of gene count distributions per sample
To get better insights into the distribution per sample, boxplots offer a good perspective.
We’ll check the distribution of read counts using a boxplot and we'll add some color to see if there is any difference between sample groups.  
```{r}
## boxplot BEFORE log2-transformation

# Check distributions of samples using boxplots
boxplot1_data <- melt(mat) %>%
  dplyr::rename("sample"="variable") %>%
  inner_join(.,groups)

p1 <- ggplot(boxplot1_data, aes(x=sample, y=value, fill=group)) + 
  geom_boxplot() + 
  theme_bw() + 
  ggtitle("Between-Sample Distribution") +
  scale_fill_manual(values=c("#6a8532", "#d95d39")) +
  ylab("Raw counts") + 
  xlab("") + 
  theme(title = element_text(size=12), 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        text = element_text(size = 10),
        axis.title=element_text(size=10),
        legend.key.size = unit(0.8, "cm")) +
  labs(fill = "Group")
p1


png(file.path(main_folder,"boxplot_pre_log-transformation.png"), width = 2500, height = 1500, units="px",res=300)
p1
dev.off()
```

```{r}
## boxplot AFTER log2-transformation

# Check distributions of samples using boxplots
boxplot2_data <- melt(log2(mat+1)) %>%
  dplyr::rename("sample"="variable") %>%
  inner_join(.,groups)
boxplot2_data$sample <- factor(boxplot2_data$sample , levels=groups$sample)

p2 <- ggplot(boxplot2_data, aes(x=sample, y=value, fill=group)) + 
  geom_boxplot() + 
  theme_bw() + 
  scale_fill_manual(values=c("#6a8532", "#d95d39")) +
  ylab("Log2 Pseudocounts") + 
  xlab("") + 
  theme(title = element_text(size=12), 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        text = element_text(size = 10),
        axis.title=element_text(size=10),
        legend.key.size = unit(0.8, "cm")) +
  labs(fill = "Group") +
  geom_hline(yintercept = median(as.matrix(log2(mat+1))), col="#297373", lwd = 0.8)
p2


png(file.path(main_folder,"boxplot_after_log-transformation.png"), width = 2500, height = 1500, units="px",res=300)
p2
dev.off()
```
From the boxplots we see that overall the density distributions of raw log-intensities are not identical but still not very different. If a sample is really far above or below the blue horizontal line we may need to investigate that sample further.



NORMALIZATION
```{r}
## boxplot before normalization
p2

## Normalization to a median of 0
# compute medians, then subtract them
channel.medians=apply(log(mat+1),2,median)
normalized.log.mat=sweep(log(mat+1),2,channel.medians,"-")

## Scale Normalization
scale.normalized.mat <- scale(normalized.log.mat)

boxplot3_data <- melt(scale.normalized.mat) %>%
  dplyr::select(-Var1) %>%
  dplyr::rename("sample"="Var2") %>%
  inner_join(.,groups)
boxplot3_data$sample <- factor(boxplot3_data$sample , levels=groups$sample)

p3 <- ggplot(boxplot3_data, aes(x=sample, y=value, fill=group)) + 
  geom_boxplot() + 
  theme_bw() + 
  scale_fill_manual(values=c("#6a8532", "#d95d39")) +
  ylab("Log2 Pseudocounts") + 
  xlab("") + 
  theme(title = element_text(size=12), 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        text = element_text(size = 10),
        axis.title=element_text(size=10),
        legend.key.size = unit(0.8, "cm")) +
  labs(fill = "Group")
p3

png(file.path(main_folder,"boxplot_after_normalization.png"), width = 2500, height = 1500, units="px",res=300)
p3
dev.off()
```

<!--
Convert Ensembl IDs to gene names using biomaRt
```{r}
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
genes <- data.frame(ensembl = rownames(mat))
# run the query
gene_IDs <- getBM(
  attributes=c("ensembl_gene_id_version","hgnc_symbol","gene_biotype"),
  filters=c("ensembl_gene_id_version"), 
  values = genes$ensembl,
  mart = ensembl)
```

count matrix is updated by matching to the retrieved annotation.
```{r}
gene_IDs <- gene_IDs %>%
  filter(!hgnc_symbol=="") 

mat_ann <- mat %>%
  rownames_to_column("ensembl_gene_id_version") %>%
  right_join(.,gene_IDs) %>%
  dplyr::select(-gene_biotype, -ensembl_gene_id_version) %>%
  group_by(hgnc_symbol) %>%
  summarise_all(mean) %>%
  column_to_rownames("hgnc_symbol")
```
-->

Mapping ensembl gene ids to hgnc symbols from GTF
```{r}
gene_IDs <- suppressMessages(read_delim(file.path(main_folder,"hgnc_conversion.tsv"))) %>%
  mutate(ensembl_gene_id_version = gsub("\\.\\d+$","",ensembl_gene_id_version,perl=TRUE))

mat_ann <- mat %>%
  rownames_to_column("ensembl_gene_id_version") %>%
  mutate(ensembl_gene_id_version = gsub("\\.\\d+$","",ensembl_gene_id_version,perl=TRUE)) %>%
  inner_join(.,gene_IDs) %>%
  dplyr::select(-gene_biotype, -ensembl_gene_id_version) %>%
  group_by(hgnc_symbol) %>%
  summarise_all(mean) %>%
  column_to_rownames("hgnc_symbol")
```

PCA on log-transformed data
```{r}
ex <- log(mat + 1)
condition <- groups$group

pca_data=prcomp(t(ex))
summary(pca_data)
screeplot(pca_data)

p4 <- fviz_eig(pca_data, 
         addlabels = TRUE, 
         ylim = c(0, 70),
         main="")
p4

png(file.path(main_folder,"pca_screeplot.png"), res=300, units = "px", width=2500, height=2000)
p4
dev.off()


pca_data_perc=round(100*pca_data$sdev^2/sum(pca_data$sdev^2),1)
df_pca_data = data.frame(PC1 = pca_data$x[,1], PC2 = pca_data$x[,2], sample = colnames(ex), condition=condition)

# PCA for components 1&2
p5 <- ggplot(df_pca_data, aes(PC1,PC2, colour = condition))+
  geom_point(size=4)+
  scale_color_manual(values=c("#6a8532", "#d95d39"))+
  labs(colour = "Group")+
  labs(x=paste0("PC1 (",pca_data_perc[1],"% variance)"), y=paste0("PC2 (",pca_data_perc[2],"% variance)") )
p5

png(file.path(main_folder,"PC1_vs_PC2.png"), res=300, units = "px", width=2500, height=2000)
p5
dev.off()



# # PCA with labels
# p4 <- ggplot(df_pca_data, aes(PC1,PC2, colour = col))+
#   geom_point(size=4, show.legend = F, colour = col)+
#   geom_label_repel(aes(label=sample, fill=condition), color="white", size=3, max.overlaps = Inf)+
#   scale_fill_manual(values=c("#6a8532", "#d95d39"))+
#   labs(x=paste0("PC1 (",pca_data_perc[1],"% variance)"), y=paste0("PC3 (",pca_data_perc[2],"% variance)") )
# 
# # Let's override some of the aesthetics
# p4 <- p4 + guides(fill = guide_legend(
#   title = "Group",
#   override.aes = aes(label = ""))
# )
# p4
# 
# png(file.path(main_folder,"PC1_vs_PC2_labels.png"), res=300, units = "px", width=2500, height=2000)
# p4
# dev.off()

# PCA for components 1&3
df_pca_data = data.frame(PC1 = pca_data$x[,1], PC3 = pca_data$x[,3], sample = colnames(ex), condition=condition)

p6 <- ggplot(df_pca_data, aes(PC1,PC3, colour = condition))+
  geom_point(size=4)+
  scale_color_manual(values=c("#6a8532", "#d95d39"))+
  labs(colour = "Group")+
  labs(x=paste0("PC1 (",pca_data_perc[1],"% variance)"), y=paste0("PC3 (",pca_data_perc[3],"% variance)") )
p6

png(file.path(main_folder,"PC1_vs_PC3.png"), res=300, units = "px", width=2500, height=2000)
p6
dev.off()
```

```{r}
ex_ann <- log(mat_ann + 1)
condition <- groups$group

pca_data=prcomp(t(ex_ann))
summary(pca_data)
screeplot(pca_data)

p4 <- fviz_eig(pca_data, 
         addlabels = TRUE, 
         ylim = c(0, 70),
         main="")
p4

png(file.path(main_folder,"pca_screeplot_2.png"), res=300, units = "px", width=2500, height=2000)
p4
dev.off()


pca_data_perc=round(100*pca_data$sdev^2/sum(pca_data$sdev^2),1)
df_pca_data = data.frame(PC1 = pca_data$x[,1], PC2 = pca_data$x[,2], sample = colnames(ex_ann), condition=condition)

# PCA for components 1&2
p5 <- ggplot(df_pca_data, aes(PC1,PC2, colour = condition))+
  geom_point(size=4)+
  scale_color_manual(values=c("#6a8532", "#d95d39"))+
  labs(colour = "Group")+
  labs(x=paste0("PC1 (",pca_data_perc[1],"% variance)"), y=paste0("PC2 (",pca_data_perc[2],"% variance)") )
p5

png(file.path(main_folder,"PC1_vs_PC2_2.png"), res=300, units = "px", width=2500, height=2000)
p5
dev.off()


# PCA for components 1&3
df_pca_data = data.frame(PC1 = pca_data$x[,1], PC3 = pca_data$x[,3], sample = colnames(ex_ann), condition=condition)

p6 <- ggplot(df_pca_data, aes(PC1,PC3, colour = condition))+
  geom_point(size=4)+
  scale_color_manual(values=c("#6a8532", "#d95d39"))+
  labs(colour = "Group")+
  labs(x=paste0("PC1 (",pca_data_perc[1],"% variance)"), y=paste0("PC3 (",pca_data_perc[3],"% variance)") )
p6

png(file.path(main_folder,"PC1_vs_PC3_2.png"), res=300, units = "px", width=2500, height=2000)
p6
dev.off()
```





```{r}
set.seed(123)
k <- 2
kmeans_result <- kmeans(t(ex_ann), k, nstart = 25)
table(kmeans_result$cluster)
```

```{r}
# Dimension reduction using PCA
res.pca <- prcomp(t(ex_ann),  scale = F)
# Coordinates of individuals
ind.coord <- as.data.frame(get_pca_ind(res.pca)$coord)
# Add clusters obtained using the K-means algorithm
ind.coord$cluster <- factor(kmeans_result$cluster)
# Add groups from the original data set
ind.coord$Group <- groups$group

# plot
p7 <- ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = c("#6a8532", "#d95d39"), ellipse = TRUE, ellipse.type = "convex",
  shape = "Group", size = 4,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("PC1 (", pca_data_perc[1], "% variance)" ),
  ylab = paste0("PC2 (", pca_data_perc[2], "% variance)" )
) +
  stat_mean(aes(color = cluster), size = 5)
p7

png(file.path(main_folder,"kmeans_2.png"), res=300, units = "px", width=2500, height=2000)
p7
dev.off()
```

```{r}
# plot
p4 <- ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "cluster", palette = c("#6a8532", "#d95d39"), ellipse = TRUE, ellipse.type = "convex",
  shape = "Group", 
  size = 4,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("PC1 (", pca_data_perc[1], "% variance)" ),
  ylab = paste0("PC2 (", pca_data_perc[2], "% variance)" )
) +
  stat_mean(aes(color = cluster), size = 8)

p4 

ggscatter(
  ind.coord, x = "Dim.1", y = "Dim.2", 
  color = "Group", palette = c("#6a8532", "#d95d39"), #ellipse = TRUE, ellipse.type = "convex",
  shape = c(16,18), 
  size = 4,  legend = "right", ggtheme = theme_bw(),
  xlab = paste0("PC1 (", pca_data_perc[1], "% variance)" ),
  ylab = paste0("PC2 (", pca_data_perc[2], "% variance)" )
) +
  stat_mean(aes(color = Group, shape = 18), size = 8)

```



Hierarchical Clustering
```{r}
# calculate distances (default: Eucledian distance)
dist_matrix <- dist(t(ex))
# perform hierarchical clustering using ? linkage 
hc1 <- hclust(dist_matrix, method = "average")
hc2 <- hclust(dist_matrix, method = "complete")
hc3 <- hclust(dist_matrix, method = "single")
hc4 <- hclust(dist_matrix, method = "ward.D2")
plot(hc1, main = "Average Linkage", hang = -1, cex = 0.6, xlab="")
plot(hc2, main = "Complete Linkage", hang = -1, cex = 0.6)
plot(hc3, main = "Single Linkage", hang = -1, cex = 0.6)
plot(hc4, main = "Ward Linkage", hang = -1, cex = 0.6)

png(file.path(main_folder,"average_linkage.png"), res=300, units = "px", width=2500, height=2000)
plot(hc1, main = "Average Linkage", hang = -1, cex = 0.6)
dev.off()
png(file.path(main_folder,"complete_linkage.png"), res=300, units = "px", width=2500, height=2000)
plot(hc2, main = "Complete Linkage", hang = -1, cex = 0.6)
dev.off()
png(file.path(main_folder,"single_linkage.png"), res=300, units = "px", width=2500, height=2000)
plot(hc3, main = "Single Linkage", hang = -1, cex = 0.6)
dev.off()
png(file.path(main_folder,"ward_linkage.png"), res=300, units = "px", width=2500, height=2000)
plot(hc4, main = "Ward Linkage", hang = -1, cex = 0.6)
dev.off()
```

```{r}
groups <- rbind(melanoma,nevi) %>%
  dplyr::select(Run,tissue) %>%
  dplyr::rename("sample"=Run, "group"=tissue)

target = c(rep(2,20), rep(1,20))

Reference = groups$group
Prediction <- cutree(hc1, k = 2)
table(Prediction,Reference)
confusionMatrix(as.factor(Prediction), as.factor(target), mode = "everything")

Reference = groups$group
Prediction <- cutree(hc2, k = 2)
table(Prediction,Reference)
confusionMatrix(as.factor(Prediction), as.factor(target), mode = "everything")

Reference = groups$group
Prediction <- cutree(hc3, k = 2)
table(Prediction,Reference)
confusionMatrix(as.factor(Prediction), as.factor(target), mode = "everything")

Reference = groups$group
Prediction <- cutree(hc4, k = 2)
table(Prediction,Reference)
confusionMatrix(as.factor(Prediction), as.factor(target), mode = "everything")
```

```{r}
Reference = groups$group
Prediction <- cutree(hc4, k = 2)
plot(table(Reference, Prediction),
     main="Confusion Matrix for Ward Clustering")
png(file.path(main_folder,"ward_linkage_confusion_matrix.png"), res=300, units = "px", width=2500, height=2000)
plot(table(Reference, Prediction), main="Confusion Matrix for Ward Clustering")
dev.off()

Prediction <- cutree(hc1, k = 2)
plot(table(Reference, Prediction),
     main="Confusion Matrix for Average Linkage")
png(file.path(main_folder,"average_linkage_confusion_matrix.png"), res=300, units = "px", width=2500, height=2000)
plot(table(Reference, Prediction), main="Confusion Matrix for Average Linkage")
dev.off()
```

```{r}
suppressPackageStartupMessages({
  library(dendextend)
})

groups <- rbind(melanoma,nevi) %>%
  dplyr::select(Run,tissue) %>%
  dplyr::rename("sample"=Run, "group"=tissue)
data <- as.data.frame(t(ex)) %>%
  rownames_to_column("sample") %>%
  inner_join(.,groups) %>%
  relocate(group, .after=sample)

groupCodes <- c(rep("primary melanoma",20), rep("benign melanocytic nevi",20))
rownames(data) <- make.unique(groupCodes)
colorCodes <- c(`primary melanoma`="#d95d39", `benign melanocytic nevi`="#6a8532")
dist_matrix <- dist(data)
hc4 <- hclust(dist_matrix, method = "ward.D2")
dend <- as.dendrogram(hc4)
# coloring the labels based on the real group category and coloring the branches based on cutting the tree into two clusters
labels_colors(dend) <- colorCodes[groupCodes][order.dendrogram(dend)]

labels(dend)
new_label <- data %>%
  dplyr::select(sample) %>%
  rownames_to_column("label")
new_label <- new_label[match(labels(dend), new_label$label),]

condition <- unique(groups$group)

dend %>%
  set("labels_cex", 0.6) %>%
  set("branches_k_color", value = c("black", "black"), k = 2) %>%
  set("labels", new_label$sample) %>%
  plot(main = "Ward Linkage")
dend %>% 
  rect.dendrogram(k=2, border = 8, lty = 1, lwd = 1.5)
legend("topright", legend = condition, fill = c("#d95d39", "#6a8532"))


png(file.path(main_folder,"ward_linkage_final.png"), res=300, units = "px", width=2500, height=2000)
dend %>%
  set("labels_cex", 0.8) %>%
  set("branches_k_color", value = c("black", "black"), k = 2) %>%
  set("labels", new_label$sample) %>%
  plot(main = "Ward Linkage") 
dend %>% 
  rect.dendrogram(k=2, border = 8, lty = 1, lwd = 1.5)
legend("topright", legend = condition, fill = c("#d95d39", "#6a8532"))
dev.off()
```

```{r}
suppressPackageStartupMessages({
  library(dendextend)
})

groups <- rbind(melanoma,nevi) %>%
  dplyr::select(Run,tissue) %>%
  dplyr::rename("sample"=Run, "group"=tissue)
data <- as.data.frame(t(ex)) %>%
  rownames_to_column("sample") %>%
  inner_join(.,groups) %>%
  relocate(group, .after=sample)

groupCodes <- c(rep("primary melanoma",20), rep("benign melanocytic nevi",20))
rownames(data) <- make.unique(groupCodes)
colorCodes <- c(`primary melanoma`="#d95d39", `benign melanocytic nevi`="#6a8532")
dist_matrix <- dist(data)
hc1 <- hclust(dist_matrix, method = "average")
dend <- as.dendrogram(hc1)
# coloring the labels based on the real group category and coloring the branches based on cutting the tree into two clusters
labels_colors(dend) <- colorCodes[groupCodes][order.dendrogram(dend)]

labels(dend)
new_label <- data %>%
  dplyr::select(sample) %>%
  rownames_to_column("label")
new_label <- new_label[match(labels(dend), new_label$label),]

condition <- unique(groups$group)

dend %>%
  set("labels_cex", 0.6) %>%
  set("branches_k_color", value = c("black", "black"), k = 2) %>%
  set("labels", new_label$sample) %>%
  plot(main = "Average Linkage")
dend %>% 
  rect.dendrogram(k=2, border = 8, lty = 1, lwd = 1.5)
legend("topright", legend = condition, fill = c("#d95d39", "#6a8532"))


png(file.path(main_folder,"average_linkage_final.png"), res=300, units = "px", width=2500, height=2000)
dend %>%
  set("labels_cex", 0.8) %>%
  set("branches_k_color", value = c("black", "black"), k = 2) %>%
  set("labels", new_label$sample) %>%
  plot(main = "Average Linkage") 
dend %>% 
  rect.dendrogram(k=2, border = 8, lty = 1, lwd = 1.5)
legend("topright", legend = condition, fill = c("#d95d39", "#6a8532"))
dev.off()
```

```{r}
Reference = groups$group
Prediction <- cutree(hc4, k = 2)
table(Prediction,Reference)

Prediction <- cutree(hc1, k = 2)
table(Prediction,Reference)
```

<!--
```{r}
##### train and test set
set.seed(1234)
nTest = floor(ncol(ex_ann) * 0.2)
ind = sample(ncol(ex_ann), nTest, FALSE)

train = ex_ann[, -ind]
dim(train)

test <- ex_ann[, ind]
dim(test)

# splitting phenodata
meta_train <- groups[-ind,]
meta_test <- groups[ind,] 

# checking if phenodata and expression data are splitted in the same way
colnames(test)
meta_test$sample
```
-->

<!--
```{r}
# build RF
rf <- randomForest(x=t(train), y=as.factor(meta_train$group), ntree=1000)
rf_pred <- predict(rf, t(test))
rf_confusion = table(rf_pred, meta_test$group)
rf_confusion

# plot of sorted importance values
varImpPlot(rf)



# selection of top 200 most important genes in RF
importance <- as.data.frame(rf$importance) %>%
  arrange(desc(MeanDecreaseGini))
top200 <- importance %>%
  top_n(200)
# write.csv(top200, file = "probes-top200.txt", quote=FALSE, row.names = FALSE, col.names=FALSE)
```
-->

<!--
```{r}
ex <- as.matrix(log(mat_ann + 1))
# ex = as.matrix(log(mat+1))
f <- factor(c(rep("melanoma",20), rep("benign melanocytic nevi",20)))

# feature selection step
tt51 <- rowttests(ex,f)
keepers <- which(tt51$p.value<0.05)
ex2 <- ex[keepers,]
tex2 <- t(ex2)
dat <- cbind(f,as.data.frame(tex2))
colnames(dat)[1] <- "GROUP"

# Split the data into training (80%) and test set (20%) 
set.seed(4321) 
training.individuals <- dat$GROUP %>%  
            createDataPartition(p = 0.8, list = FALSE) 
train <- as.vector(training.individuals)
test <- setdiff(1:40, train)
length(train)
length(test)

mod <- lda(GROUP ~ ., data=dat, subset = train)

mod.values <- predict(mod, dat[train,])
mod.values$class
ldahist(data = mod.values$x[,1], g = dat[train, "GROUP"])
png(file.path(main_folder,"lda_histogram.png"), res=300, units = "px", width=2500, height=2000)
ldahist(data = mod.values$x[,1], g = dat[train, "GROUP"])
dev.off()

plot(mod.values$x[,1], ylab=c("LDA Axis"))
text(mod.values$x[,1],
col=c(as.numeric(dat[train,"GROUP"])+10))
png(file.path(main_folder,"lda_plot.png"), res=300, units = "px", width=2500, height=2000)
plot(mod.values$x[,1], ylab=c("LDA Axis"))
text(mod.values$x[,1],
col=c(as.numeric(dat[train,"GROUP"])+10))
dev.off()

preds <- predict(mod, dat[test,])
preds$class
table(as.numeric(preds$class),
as.numeric(dat[test, "GROUP"]) )
# ROC
roc_lda <- plot.roc(as.numeric(preds$class),
as.numeric(dat[test, "GROUP"]) )
png(file.path(main_folder,"roc_lda.png"), res=300, units = "px", width=2500, height=2000)
plot.roc(as.numeric(preds$class),
as.numeric(dat[test, "GROUP"]) )
dev.off()
```
-->

```{r}
ex <- as.matrix(log(mat_ann + 1))
f <- factor(c(rep("melanoma",20), rep("benign melanocytic nevi",20)))

# feature selection step
tt51 <- rowttests(ex,f)
keepers <- which(tt51$p.value<0.05)
ex2 <- ex[keepers,]
tex2 <- t(ex2)
dat <- cbind(f,as.data.frame(tex2))
colnames(dat)[1] <- "GROUP"

#dat <- cbind(f,as.data.frame(t(ex)))
#colnames(dat)[1] <- "GROUP"

# Split the data into training (80%) and test set (20%) 
set.seed(4321) 
training.individuals <- dat$GROUP %>%  
            createDataPartition(p = 0.8, list = FALSE) 
train <- as.vector(training.individuals)
test <- setdiff(1:40, train)
length(train)
length(test)
```

<!--
```{r}
test_set <- read.csv(file.path(main_folder,"additional_test_set.tsv"), header = TRUE, row.names = 1, sep = "\t", check.names = FALSE)
```
-->

```{r}
# Run algorithms using 10-fold cross validation, 10 times
control <- trainControl(method="cv", number=10, repeats = 10)
metric <- "Accuracy"

# LDA
set.seed(1234)
fit.lda <- train(GROUP~., data=dat[train,], method="lda", metric=metric, trControl=control)
fit.lda


# RF
set.seed(1234)
fit.rf <- train(GROUP~., data=dat[train,], method="rf", metric=metric, trControl=control, ntree = 1000)
fit.rf
fit.rf$results
fit.rf$finalModel
# feature importances
varImp(fit.rf)
rf_genes <- as.data.frame(fit.rf$finalModel$importance) %>%
  arrange(-MeanDecreaseGini) %>%
  rownames_to_column("gene") %>%
  mutate(gene=gsub("`","",gene))
write.table(rf_genes$gene, file.path(main_folder,"rf_genes.txt"), sep = "\t", quote=FALSE, row.names=F, col.names=F) 

randomForest::varImpPlot(fit.rf$finalModel, main="Variable Importance") 
png(file.path(main_folder,"rf_importance.png"), res=300, units = "px", width=2500, height=2000)
randomForest::varImpPlot(fit.rf$finalModel, main="Variable Importance")
dev.off()


# LASSO
set.seed(1234)
fit.lasso <- train(GROUP~., data=dat[train,], 
                   method="glmnet", trControl = control, metric = metric,
                   family = "binomial", 
                   tuneGrid = expand.grid(alpha = 1, lambda = seq(0.01,3, by=0.01))
                   )
plot(fit.lasso)



preds.lda <- predict(fit.lda, dat[test,])
table(preds.lda, dat[test,]$GROUP)
confusionMatrix(preds.lda, dat[test,]$GROUP, mode = "everything") 

test_set$GROUP <- factor(test_set$GROUP, levels = c("benign melanocytic nevi", "melanoma"))
preds.lda <- predict(fit.lda, test_set)
table(preds.lda, test_set$GROUP)
confusionMatrix(preds.lda, test_set$GROUP, mode = "everything")

preds.rf <- predict(fit.rf, dat[test,])
table(preds.rf, dat[test,]$GROUP)
confusionMatrix(preds.rf, dat[test,]$GROUP, mode = "everything") 

preds.rf <- predict(fit.rf, test_set)
table(preds.rf, test_set$GROUP)
confusionMatrix(preds.rf, test_set$GROUP, mode = "everything")
```

```{r}
# LASSO
set.seed(1234) 
fit.lasso0 <- train(GROUP~., data=dat[train,], 
                   method="glmnet", trControl = control, metric = metric,
                   family = "binomial", 
                   tuneGrid = expand.grid(alpha = 1, lambda = seq(0.01,3, by=0.01))
                   )

# Best tuning parameter to minimize the root mean squared error (RMSE) of model
fit.lasso0$bestTune
fit.lasso0$bestTune$lambda
print(paste0('Lasso best parameters: ' , fit.lasso0$finalModel$lambdaOpt))
plot(fit.lasso0)
png(file.path(main_folder,"lasso0_lambda.png"), res=300, units = "px", width=2500, height=2000)
plot(fit.lasso0)
dev.off()

# Re-train the model with the a new grid with smaller values of lambda
fit.lasso <- train(GROUP~., data=dat[train,], 
                   method="glmnet", trControl = control, metric = metric,
                   family = "binomial", 
                   tuneGrid = expand.grid(alpha = 1, lambda = seq(0.01,0.5, by=0.001))
                   )
fit.lasso$bestTune
fit.lasso$bestTune$lambda
print(paste0('Lasso best parameters: ' , fit.lasso$finalModel$lambdaOpt))
plot(fit.lasso)
png(file.path(main_folder,"lasso_lambda.png"), res=300, units = "px", width=2500, height=2000)
plot(fit.lasso)
dev.off()

fit.lasso$results[124,]
```
```{r}
# LASSO regression model coefficients (parameter estimates)
# Regression Coefficients for the final model (with the best performance)
coefs <- coef(fit.lasso$finalModel, # Select the final model coefficients
     fit.lasso$bestTune$lambda) # at the best lambda value
coefs
coefs.zero <- coefs[which(coefs[,1]==0),]
length(coefs.zero)

# Rank order the predictors based on the magnitude of the coefficients
ind <- order(abs(coefs),decreasing=T)
lasso_genes <- data.frame(coef = coefs[ind[-1],]) %>%
  rownames_to_column("gene") %>%
  filter(coef != 0) %>%
  mutate(gene=gsub("`","",gene))
write.table(lasso_genes$gene, file.path(main_folder,"lasso_genes.txt"), sep = "\t", quote=FALSE, row.names=F, col.names=F) 


# Variable importance
suppressPackageStartupMessages({
  library(vip)
})

vip(fit.lasso, 
    num_features = 12, 
    geom = "point") + 
theme_bw()
png(file.path(main_folder,"lasso_importance_plot.png"), res=300, units = "px", width=2500, height=2000)
vip(fit.lasso, 
    num_features = 12, 
    geom = "point") + 
theme_bw()
dev.off()
```

```{r}
# plot coefficient paths as a function of lambda
# plot(fit.lasso$finalModel, xvar = "lambda", label=TRUE, col = rainbow(20))
plot(fit.lasso$finalModel, xvar = "lambda", label=FALSE, col = rainbow(20))
abline(v=-2.017406)
# log(fit.lasso$bestTune$lambda)
png(file.path(main_folder,"lasso_coeff.png"), res=300, units = "px", width=2500, height=2000)
plot(fit.lasso$finalModel, xvar = "lambda", label=FALSE, col = rainbow(20))
abline(v=-2.017406)
dev.off()

# codebook for which variables the numbers correspond to
rownames(fit.lasso$finalModel$beta)

# e.g., What are variables 519 and 4?
rownames(fit.lasso$finalModel$beta)[c(419,4)]
```

```{r}
### Model prediction
# Goal: See how well our model predicts when we give it new data
predictions_LASSO <- predict(fit.lasso, # Use trained model
                             newdata = dat[test,]) # To predict outcome with test data
confusionMatrix(predictions_LASSO,dat[test,]$GROUP, mode = "everything")

test_set$GROUP = factor(test_set$GROUP, levels=c("benign melanocytic nevi", "melanoma"))
predictions_LASSO <- predict(fit.lasso,
                             newdata = test_set)
confusionMatrix(predictions_LASSO,test_set$GROUP, mode = "everything")


preds.lasso2 <- predict(fit.lasso, dat[test,], type="raw", s=fit.lasso$finalModel$lambdaOpt)
table(preds.lasso2, dat[test,]$GROUP)
preds.lasso <- predict(fit.lasso, dat[test,], type="prob")

f.categorical <- c(rep("primary melanoma",20), rep("benign melanocytic nevi",20))
auc <- roc(f.categorical[test], preds.lasso[[2]])
print(auc$auc)
plot(auc) #area under the curve 0.9375

png(file.path(main_folder,"lasso_auc.png"), res=300, units = "px", width=2500, height=2000)
plot(auc)
dev.off()
```

```{r}
# comparison with other classification methods
results <- resamples(list(RF=fit.rf, LDA=fit.lda, Lasso=fit.lasso))
summary(results)
ggplot(results) + labs(y = "Accuracy") + theme_linedraw()
```

<!--
```{r}
# rScudo
y <- c(rep(0,20),rep(1,20))
f <- factor(y, labels = c("melanoma","benign melanocytic nevi"))
library(rScudo)
trainData <- t(tex2[train,])
testData <- t(tex2[test,])

# analyze training set
trainRes <- scudoTrain(trainData, groups = dat$GROUP[train], nTop = 10, nBottom = 10, alpha = 0.05)
trainRes
# inspect signatures
upSignatures(trainRes)[1:5,1:5]
consensusUpSignatures(trainRes)[1:5, ]
# generate and plot map of training samples
trainNet <- scudoNetwork(trainRes, N = 0.2)
scudoPlot(trainNet, vertex.label = NA)

# perform validation using testing samples
testRes <- scudoTest(trainRes, testData, f[test], nTop = 20, nBottom = 20)
testNet <- scudoNetwork(testRes, N = 0.2)
scudoPlot(testNet, vertex.label = NA)

# identify clusters on map
library("igraph")
testClust <- igraph::cluster_spinglass(testNet, spins = 2)
plot(testClust, testNet, vertex.label = NA)
# perform classification
classRes <- scudoClassify(trainData, testData, N = 0.25, nTop = 12, nBottom = 12, trainGroups = f[inTrain], alpha = 0.5)
caret::confusionMatrix(classRes$predicted, f[-inTrain])
```
-->

```{r}
# rScudo
y <- c(rep(0,20),rep(1,20))
f <- factor(y, labels = c("melanoma","benign melanocytic nevi"))
library(rScudo)
trainData <- t(tex2[train,])
testData <- t(tex2[test,])


set.seed(1234)
# use caret to test a grid a values for nTop & nBottom using cross validation
model <- scudoModel(nTop = (1:6)*5, nBottom = (1:6)*5, N = 0.25)
control <- caret::trainControl(method = "cv", number = 5, summaryFunction = caret::multiClassSummary)
cvRes <- caret::train(x = t(trainData), y = dat$GROUP[train], method = model, trControl = control)

trainRes <- scudoTrain(trainData, groups = dat$GROUP[train], nTop = cvRes$bestTune$nTop, nBottom = cvRes$bestTune$nBottom, alpha = 0.05)
trainRes
# generate and plot map of training samples
set.seed(1234)
trainNet <- scudoNetwork(trainRes, N = 0.35)
library("igraph")
igraph::V(trainNet)$color
V(trainNet)$color <- ifelse(V(trainNet)$color == "#00FFFF", "#d95d39", "#6a8532")
scudoPlot(trainNet, vertex.label = NA, x = "bottom")

png(file.path(main_folder,"scudo_train.png"), res=300, units = "px", width=2500, height=2000)
scudoPlot(trainNet, vertex.label = NA, x = NA)
dev.off()


#library(RCy3)
#scudoCytoscape(trainNet, title = "Scudo Graph", collection = "SCUDO")

# identify clusters on map
trainClust <- igraph::cluster_spinglass(trainNet, spins = 2)
plot(trainClust, trainNet, vertex.label = NA)
png(file.path(main_folder,"scudo_train_clusters.png"), res=300, units = "px", width=2500, height=2000)
plot(trainClust, trainNet, vertex.label = NA)
dev.off()
```

```{r}
# plot map of testing samples using best nTop & nBottom values
set.seed(1234)
testRes <- scudoTest(trainRes, testData, f[test], cvRes$bestTune$nTop, cvRes$bestTune$nBottom)
testNet <- scudoNetwork(testRes, N = 0.3)
scudoPlot(testNet, vertex.label = NA)

# identify clusters on map
library("igraph")
testClust <- igraph::cluster_spinglass(testNet, spins = 2)
plot(testClust, testNet, vertex.label = NA)
# perform classification of testing samples using best nTop & nBottom values
classRes <- scudoClassify(trainData, testData, N = 0.25, cvRes$bestTune$nTop, cvRes$bestTune$nBottom, trainGroups = f[train], alpha = 0.5)
caret::confusionMatrix(classRes$predicted, f[test])
```

<!--
```{r}
testData <- t(test_set %>%
                dplyr::select(-GROUP))
y <- c(rep(0,3),rep(1,37))
f <- factor(y, labels = c("benign melanocytic nevi", "melanoma"))
set.seed(1234)
testRes <- scudoTest(trainRes, testData, f, cvRes$bestTune$nTop, cvRes$bestTune$nBottom)
testNet <- scudoNetwork(testRes, N = 0.3)
scudoPlot(testNet, vertex.label = NA)

# identify clusters on map
testClust <- igraph::cluster_spinglass(testNet, spins = 2)
plot(testClust, testNet, vertex.label = NA)
# perform classification of testing samples using best nTop & nBottom values
classRes <- scudoClassify(trainData, testData, N = 0.25, cvRes$bestTune$nTop, cvRes$bestTune$nBottom, trainGroups = f[train], alpha = 0.5)
caret::confusionMatrix(classRes$predicted, f, mode = "everything")
```
-->

extract signatures
```{r}
scudo_up <- trainRes@consensusUpSignatures
scudo_down <- trainRes@consensusDownSignatures

scudo_genes <- rbind(scudo_up,scudo_down)

write.table(sort(scudo_up$melanoma), file.path(main_folder,"scudo_up_melanoma.txt"), sep = "\t", quote=FALSE, row.names=F, col.names=F)
write.table(sort(scudo_down$melanoma), file.path(main_folder,"scudo_down_melanoma.txt"), sep = "\t", quote=FALSE, row.names=F, col.names=F)

write.table(sort(scudo_up$`benign melanocytic nevi`), file.path(main_folder,"scudo_up_nevi.txt"), sep = "\t", quote=FALSE, row.names=F, col.names=F)
write.table(sort(scudo_down$`benign melanocytic nevi`), file.path(main_folder,"scudo_down_nevi.txt"), sep = "\t", quote=FALSE, row.names=F, col.names=F)
```



```{r}
summary <- data.frame("Classification Algorithm"= c("RF","LDA","LASSO","SCUDO"),
                      "Training Accuracy Score"= c("80.8%","84.1%","87.5%","84.3%"),
                      "Test Accuracy Score"= c("100%","100%","87.5%","100%"),
                      "2nd Test Accuracy Score"= c("87.5%","87.5%","92.5%","92.5%")
                      )
colnames(summary) <- c("Classification Algorithm","Training Accuracy Score","Test Accuracy Score", "2nd Test Accuracy Score")


summary2 <- data.frame("Classification Algorithm"= c("RF","LDA","LASSO","SCUDO"),
                      "Training Accuracy Score"= c("80.8%","84.1%","87.5%","84.3%"),
                      "Test Accuracy Score"= c("100%","100%","87.5%","100%"),
                      "2nd Test Accuracy Score"= c("87.5%","87.5%","92.5%","92.5%"),
                      "2nd Test F1 Score"= c("0.5455","0.5455","0.5714","0")
                      )
colnames(summary2) <- c("Classification Algorithm","Training Accuracy Score","Test Accuracy Score", "2nd Test Accuracy Score", "2nd Test F1 Score")
```


```{r}
summary <- data.frame("Classification Algorithm"= c("RF","LDA","LASSO","SCUDO"),
                      "Training Accuracy Score"= c("82.5%","82.5%","100%","87.1%"),
                      "Test Accuracy Score"= c("100%","100%","87.5%","100%"),
                      "2nd Test Accuracy Score"= c("42.5%","50.0%","45.0%","95.0%")
                      )
colnames(summary) <- c("Classification Algorithm","Training Accuracy Score","Test Accuracy Score", "2nd Test Accuracy Score")


summary2 <- data.frame("Classification Algorithm"= c("RF","LDA","LASSO","SCUDO"),
                      "Training Accuracy Score"= c("82.5%","82.5%","100%","87.1%"),
                      "Test Accuracy Score"= c("100%","100%","87.5%","100%"),
                      "2nd Test Accuracy Score"= c("42.5%","50.0%","45.0%","95.0%"),
                      "2nd Test F1 Score"= c("0","0.16667","0","0.5000")
                      )
colnames(summary2) <- c("Classification Algorithm","Training Accuracy Score","Test Accuracy Score", "2nd Test Accuracy Score", "2nd Test F1 Score")
```



AUTOGO
```{r}
suppressPackageStartupMessages({
  library(DESeq2)
  library(xlsx)
})
```

```{r}
files <- list.files(path = "/home/alice/ago2023/Desktop/auto-go/R", recursive = T, all.files = T, full.names = T)
invisible(sapply(files, source))

# main_folder <- "/home/alice/ago2023/Desktop/unitn/NDA/autogo/"
```

```{r}
counts <- mat_ann %>%
  rownames_to_column("gene_id")

comparisons <- data.frame(treatment = "primary melanoma",
                          control = "benign melanocytic nevi")
```

```{r}
# deseq_analysis(counts, groups, comparisons, padj_threshold = 0.05, log2FC_threshold = 1, pre_filtering = T, save_excel = T, where_results = main_folder, outfolder = "differential_analysis_0724/")

suppressPackageStartupMessages({
  library(pathfindR)
})

autogo <- readxl::read_xlsx(file.path(main_folder,"autogo/DE_benign melanocytic nevi_vs_primary melanoma_allres.xlsx"))
scudo_all <- read_tsv(file.path(main_folder,"scudo_genes.txt"), show_col_types = FALSE, col_names = "gene")
all_scudo_genes <- autogo %>%
  filter(genes %in% scudo_all$gene) %>%
  dplyr::select(genes,log2FoldChange,padj)
all_scudo_genes <- as.data.frame(all_scudo_genes)
  
output_df1 <- run_pathfindR(all_scudo_genes, iterations = 1, gene_sets = "GO-MF", min_gset_size = 5, output_dir = "/home/alice/ago2023/Desktop/unitn/NDA/pathfindr-GO-MF")

output_df2 <- run_pathfindR(all_scudo_genes, iterations = 1, gene_sets = "KEGG", min_gset_size = 5, output_dir = "/home/alice/ago2023/Desktop/unitn/NDA/pathfindr-KEGG")

output_df3 <- run_pathfindR(all_scudo_genes, iterations = 1, gene_sets = "GO-All", min_gset_size = 5, output_dir = "/home/alice/ago2023/Desktop/unitn/NDA/pathfindr-GO-All")

## cluster enriched terms
cluster_enriched_terms(output_df3)
## term-gene graph of top 10 terms
term_gene_graph(output_df3)
## visualizing enriched term diagrams
visualize_terms(output_df3)

term_gene_heatmap(output_df3)
```




```{r}
up_melanoma <- read_tsv(file.path(main_folder,"scudo_up_melanoma.txt"), show_col_types = FALSE, col_names = "gene")
down_melanoma <- read_tsv(file.path(main_folder,"scudo_down_melanoma.txt"), show_col_types = FALSE, col_names = "gene")
```




